On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-09-12 Manuscript/manuscript.tex
Edited a few tables, fixed spacing in data section

Devon Allen, an alumnus of the University of Oregon in 
Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field 
Championships also in Eugene in late June, Allen was poised to compete in the 
2022 World Track and Field Championships at the same Eugene venue in August.
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 and 0.101 in his preliminary heat and semifinal, respectively.  
His disqualification in the final heat by reacting in 0.099 seconds, just 0.001
seconds faster than the 0.1 threshold set by the International Association of
Athletics Federations (IAAF), was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification
(\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}).
The crowd erupted with jeers
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting until the next World Track and Field Championships.
\eds{He is not competing this year, correct? If yes, we may need to rephrase.}
\of{Correct}

@@

There are two types of data that are explored throughout this paper: data from
1999-2023 for every world championship, and data from athletes who competed in
the 2022 World Championships and another competition (either 2019 World 
Championships, 2023 World Championships or a 2022 national competition typically
held within three months of the 2022 World Championships). 
%\eds{is this true? are you still using the 2019 world championships data for the
%rank-based analyses, or only the national competitions?}
%\of{Yes the ranked based comparison has 2 subsections: 2022 national vs 2022
%international and 2019 international vs 2022 international.  Should I instead say
%that there are three types of data across 2 sections or something like that?}
%\eds{I think this is fine here, but it is not at all clear that there are two
%subsections in the rank-based results for 2022 national vs 2022
%international and 2019 international vs 2022 international.}
%\of{Currently under the Methods/Results section I have GLMM and rank based comparison
%do you think I need subsections under rank based comparison, or is that too much
%compartmentalization?  Or should I instead say there are 3 types of data that are
%used in two different methods, or something like that?} \eds{I prefer
%subsections within rank-based comparison.}
The first data set is used for the GLMM analysis described in 
Section~\ref{sec:glmm}, while the second collection of data is used for the 
rank-based comparisons described in Section~\ref{sec:rank}.  This paper was
started shortly after the 2022 World Championships, but as the 2023 data has 
become available recently, we are pleased to include it in our analysis, and it
did not significantly change our results.

@@

\subsubsection{2022 National Competitions}\label{sec:datanational}
From June 23-26, 2022, the United States held its Track and Field Championships 
at Hayward Field in Eugene, Oregon to decide who to send to the World 
Championships being held at the same venue in August. Thus, we can establish a 
baseline for the four United States athletes who competed in at least one round 
of both events: Trey Cunningham, Daniel Roberts, Grant Holloway, and Devon Allen.   
Every athlete reacted faster in all the World Track and Field Championships 
races compared to the United States Track and Field Championships races. The 
dataset of the four United States athletes is too small to perform any 
meaningful statistical analysis, but the data was expanded to include reaction 
times for 11 more athletes who competed at other national competitions (many of which were 
held from May-July of 2022). This data was difficult to find, as the data is not
centrally located but instead found on different national websites (often 
recorded in the national language without translation, and hence not used). 
\of{I don't know how to say this}. \eds{is my change correct?} \of{No I was able
to use a lot even though I couldn't actually read the page. I just tried to find
110 meter hurdle and then look at a column for reaction times which are always from
roughly 0.1 to 0.18 seconds.} Each athlete was 
designated as a ``cluster'' and the stage of competition (national or international) 
was specified for each competition within the cluster (athlete).
Thus, this is subunit level grouping where we assign the treatment, national or
international, within each cluster.  Cluster sizes range from three to six, with
one cluster of size three, six clusters of size four, two clusters of size five,
and two clusters of size six.  We chose to include preliminary rounds in this
section as well as in Section~\ref{sec:data2019} in order to increase the size 
of our data.  Without data from the preliminary rounds, the number of observations
dropped from 65 to 26 and the number of athletes included in the analysis
decreased from 15 to 9.

@@

\begin{table}
  \centering
  \caption{Suggested reaction time barriers based on probability of observing
  a time less than the barrier. \of{Not sure if this makes sense}}
  \begin{tabular}{c c c} 
   \toprule
   Data Set & P(RxnTime $< x = 10^{-3}$) & P(RxnTime $< x = 10^{-4}$) \\ 
   \midrule
   Excluding 2022 & $0.094$ & $0.084$ \\ 
   Including 2022 & $0.092$ & $0.083$ \\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_time}
\end{table}

@@

\begin{table}
  \centering
  \caption{}
  \begin{tabular}{c c c c c} 
   \toprule
   Comparison & Permutation & Asymptotic \\ 
   \midrule
   2019 vs 2022 & $2.9\cdot10^{-3}$ & $2.3\cdot10^{-2}$ \\
   2022 vs 2023 & $3.0\cdot10^{-3}$ & $1.0\cdot10^{-2}$ \\
   2022 National vs International & $2.2\cdot10^{-16}$ & $1.0\cdot10^{-3}$ \\
   \bottomrule
  \end{tabular}
  \caption{2019 vs 2022 compares data from the same athletes who competed at the
  2019 and 2022 World Track and Field Championships. 2022 vs 2023 compares data
  from the same athletes who competed at the 2022 and 2023 World Track and Field
  Championships. Nat vs Inter compares data from the same athletes who competed
  at the 2022 national-level championships and the 2022 World Track and Field
  Championships.}
  \label{tab:Clusrankresults}
\end{table}

@@

>>>>

Devon Allen, an alumnus of the University of Oregon in Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field Championships also in Eugene 
in late June, Allen was poised to compete in the 2022 World Track and 
Field Championships in August, which would also be held in Eugene, Oregon.
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 and 0.101 in his preliminary heat and semifinal, respectively.  
His disqualification in the final heat by reacting in 0.099 seconds, just 0.001
seconds faster than the 0.1 threshold set by the International Association of
Athletics Federations (IAAF), was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification
(\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}).
The crowd erupted with jeers
when they found out that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete and been 
spared waiting until the next World Track and Field Championships.
\eds{He is not competing this year, correct? If yes, we may need to rephrase.}
\of{Correct}

@@

There are two types of data that are explored throughout this paper: data from
1999-2023 for every world championship, and data from athletes who competed in
the 2022 World Championships and another competition (either 2019 World 
Championships, 2023 World Championships or a 2022 national competition typically
held within three months of the 2022 World Championships).

The first data set is used for the GLMM analysis described in 
Section~\ref{sec:glmm}, while the second collection of data is used for the 
rank-based comparisons described in Section~\ref{sec:rank}.  This paper was
started shortly after the 2022 World Championships, but as the 2023 data has 
become available recently, we are pleased to include it in our analysis, and it
did not significantly change our results.

@@

\subsubsection{2022 National Competitions}\label{sec:datanational}
From June 23-26, 2022, the United States held its Track and Field Championships 
at Hayward Field in Eugene, Oregon to decide who to send to the World 
Championships being held at the same venue in August. Thus, we can establish a 
baseline for the four United States athletes who competed in at least one round 
of both events: Trey Cunningham, Daniel Roberts, Grant Holloway, and Devon Allen.   
Every athlete reacted faster in all the World Track and Field Championships 
races compared to the United States Track and Field Championships races. The 
dataset of the four United States athletes is too small to perform any 
meaningful statistical analysis, but the data was expanded to include reaction 
times for 11 more athletes who competed at other national competitions (many of which were 
held from May-July of 2022). This data was difficult to find, as the data is not
centrally located but instead found on different national websites (often 
recorded in the national language without translation, and hence not used). 
\of{I don't know how to say this}. \eds{is my change correct?} \of{No I was able
to use a lot even though I couldn't actually read the page. I just tried to find
110 meter hurdle and then look at a column for reaction times which are always from
roughly 0.1 to 0.18 seconds.}

Each athlete was designated as a ``cluster'' and the stage of competition (national
or international) was specified for each competition within the cluster (athlete).
Thus, this is subunit level grouping where we assign the treatment, national or
international, within each cluster.  Cluster sizes range from three to six, with
one cluster of size three, six clusters of size four, two clusters of size five,
and two clusters of size six.  We chose to include preliminary rounds in this
section as well as in Section~\ref{sec:data2019} in order to increase the size 
of our data.  Without data from the preliminary rounds, the number of observations
dropped from 65 to 26 and the number of athletes included in the analysis
decreased from 15 to 9.

@@

\begin{table}
  \centering
  \caption{Suggested reaction time barriers based on probability of observing
  a time less than the barrier. \of{Not sure if this makes sense}}
  \begin{tabular}{c c c} 
   \toprule
   Data Set & P(RxnTime $< x) = 10^{-3}$ & P(RxnTime $< x) = 10^{-4}$ \\ 
   \midrule
   Excluding 2022 & $0.094$ & $0.084$ \\ 
   Including 2022 & $0.092$ & $0.083$ \\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_time}
\end{table}

@@

\begin{table}
  \centering
  \caption{2019 vs 2022 compares data from the same athletes who competed at the
  2019 and 2022 World Track and Field Championships. 2022 vs 2023 compares data
  from the same athletes who competed at the 2022 and 2023 World Track and Field
  Championships. Nat vs Inter compares data from the same athletes who competed
  at the 2022 national-level championships and the 2022 World Track and Field
  Championships.}
  \begin{tabular}{c c c c c} 
   \toprule
   Comparison & Permutation & Asymptotic \\ 
   \midrule
   2019 vs 2022 & $2.9\cdot10^{-3}$ & $2.3\cdot10^{-2}$ \\
   2022 vs 2023 & $3.0\cdot10^{-3}$ & $1.0\cdot10^{-2}$ \\
   2022 National vs International & $2.2\cdot10^{-16}$ & $1.0\cdot10^{-3}$ \\
   \bottomrule
  \end{tabular}
  \label{tab:Clusrankresults}
\end{table}

@@

This paper has shown that a reaction time of 0.1 seconds is not as extreme as
one may think, and that based on the probabilities in table~\ref{tab:Sim_time},
World Athletics should consider changing the disqualification barrier to 0.8
seconds.  If they were to do so, they would allow athletes like Allen to be able
to react faster without fear of being disqualified.  The second part of this
paper showed that athletes' reaction times at the 2022 World Track and Field
Championships were generally faster than their reaction times at other
competitions.
